---
title: "Power Analysis for Conflict Designs"
author: "Matthew Crump"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(conflictPower)
```


### c_score()

The c_score function generates a single simulated conflict score based on samples from a specified RT distribution for conflict and no-conflict conditions. The examples below defines the conflict effect as a difference of 100 between the mu parameters, and assumes 10 trials per condition.

The c_score function samples the specified number of RTs from each distribution, computes the mean for each condition (conflict and no conflict), and returns the difference (conflict - no conflict).

```{r}
 c_score(c_nmst= c(n=10, mu=600, sigma=100, tau=100),
         nc_nmst= c(n=10, mu=500, sigma=100, tau=100))
```

### c_scores()

`c_scores()` generates n number of conflict scores. This can be useful for simulating n number of subjects in an experiment, or for simulating a sampling distribution of conflictd scores.

The sampling distribution is useful for understanding the precision of the design for measuring a particular individual's conflict score. For example, with only 10 trials in each condition, the distribution of possible conflict scores for this subject is very wide. 

```{r}
x <- c_scores(subjects = 1000,
            c_nmst= c(n=10, 
                      mu=600, 
                      sigma=100, 
                      tau=100),
            nc_nmst= c(n=10, 
                       mu=500, 
                       sigma=100, 
                       tau=100))
mean(x)
sd(x)
hist(x)
```

### c_power()

This function computes statistical power to detect a conflict effect using monte carlo simulation. 

```{r}
c_power(subjects=20,
             c_nmst = c(20,500,100,100),
             nc_nmst = c(20,550,100,100),
             num_sims = 1000,
             alpha = .05)
```

